\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-02-07 papers/methods/paper.tex
Slight paper update

In this paper, we lay out a new hierarchical model that advances the past work of \citet{Rubin2015}. Section \ref{sec:review} is dedicated to a quick review of the supernovae cosmology. In Section \ref{sec:method} we outline our methodology. Model verification on simulated datasets is given in Section \ref{sec:verification}. Forecasts for the impending DES three year spectroscopic supernova survey are contained in Section \ref{sec:forecasts}. {\red Section \ref{sec:sys} investigates the effect of various systematics on our model, and Section \ref{sec:details} provides details on potential areas of improvement and unsuccessful methodologies. }

@@

Traditional $\chi^2$ analyses such as that found in \citet{Kowalski2008, Conley2011, Betoule2014}, minimise the difference in distance modulus between the cosmologically predicted values $\mu_C$ and the observed distance modulus $\mu_{\rm obs}$, shown respectively below:
\begin{align}
\mu_C &= 5 \log\left[ \frac{(1+z)r}{10} \right]\\
r& =\frac{c}{H_0} \int_0^z \frac{dz'}{\sqrt{ \Omega_m (1+z')^3 + \Omega_k (1 + z')^2 + \Omega_\Lambda (1+z')^{3(1+w)}}} \label{eq:mu0}\\
\mu_{\rm obs} &= m_B + \alpha x_1 - \beta c - M_B \label{eq:mu1}
\end{align}
The minimising function is then given as
\begin{equation}
\chi^2 = (\mu_{\rm obs} - \mu_C)^\dagger C^{-1} (\mu_{\rm obs} - \mu_C)
\end{equation}
where $C^{-1}$ is an uncertainty matrix that combined the uncertainty from the SALT2 fits, intrinsic dispersion, calibration, dust, peculiar velocity and many other factors (see \citet{Betoule2014} for a review). The benefit this analysis methodology provides is speed - for samples of hundreds of supernovae, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with two costs. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of uncertainty Gaussianity. Secondly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae.

Selection efficiency, such as the well known Malmquist bias \citep{MalmquistK.G.1922} is accounted for by correcting data. Simulations following survey observational strategies and geometry are used to calculate the expected bias in distance modulus, which is then added onto the observational data. As these effects are not built into the likelihood, their influence on the error budget is not captured fully in the $\chi^2$ distribution.

@@

Sitting comfortably between the traditional models simplicity and the complexity of forward modelling lies Hierarchical Bayesian Models. With the introduction of multiple layers in our model, we can add far more flexibility than a traditional analysis whilst still maintaining most of the computational benefits that come from having a tractable likelihood. \citet{Mandel2009} and \citet{Mandel2011} construct a hierarchical model that they apply to the light curve fitting for supernova. \citet{March2011, March2014, Karpenka2015} derive a hierarchical model and simplify it by analytically marginalising over nuisance parameters to provide a model that offers increased flexibility with reduced uncertainty over the traditional method. The recent BAHAMAS model \citep{Shariff2016} builds off this and reanalyses the JLA dataset, whilst including extra freedom in the correction factors $\alpha$ and $\beta$, finding evidence for redshift dependence on $\beta$. \citet{Ma2016} also performed a reanalysis of the JLA dataset, finding significant differences in $\alpha$ and $\beta$ values from the original as well. Notably, these methods rely on data that is bias corrected, however the UNITY framework given by \citet{Rubin2015} incorporates selection efficiency analytically in the model, and is applied to the Union 2.1 dataset \citep{Suzuki2012}. The well known BEAMS (Bayesian estimation applied to multiple species) methodology from \citet{Kunz2007} has been extended and applied in several works \citep{Hlozek2012}, mostly lately to include redshift uncertainty for photometric redshift application as zBEAMS \citep{Roberts2017} and to include simulated bias corrections in \citet{Kessler2017}.

The flexibility afforded by a hierarchical model allows for investigations into different treatments of underlying populations, rates, redshift distributions, mass corrections and redshift evolution, each of will will be discussed further in the outline of our model below.

>>>>

In this paper, we lay out a new hierarchical model that advances the past work of \citet{Rubin2015} to include more robust treatment of systematics and selection effects whilst improving computational performance. Section \ref{sec:review} is dedicated to a quick review of the supernovae cosmology. In Section \ref{sec:method} we outline our methodology. Model verification on simulated datasets is given in Section \ref{sec:verification}. Forecasts for the impending DES three year spectroscopic supernova survey are contained in Section \ref{sec:forecasts}. {\red Section \ref{sec:sys} investigates the effect of various systematics on our model, and Section \ref{sec:details} provides details on potential areas of improvement and unsuccessful methodologies. }

@@

Traditional $\chi^2$ analyses such as that found in \citet{Kowalski2008, Conley2011, Betoule2014}, minimise the difference in distance modulus between the cosmologically predicted values $\mu_C$ and the observed distance modulus $\mu_{\rm obs}$, shown respectively below:
\begin{align}
\mu_C &= 5 \log\left[ \frac{(1+z)r}{10} \right]\\
r& =\frac{c}{H_0} \int_0^z \frac{dz'}{\sqrt{ \Omega_m (1+z')^3 + \Omega_k (1 + z')^2 + \Omega_\Lambda (1+z')^{3(1+w)}}} \label{eq:mu0}\\
\mu_{\rm obs} &= m_B + \alpha x_1 - \beta c - M_B \label{eq:mu1}
\end{align}
The minimising function is then given as
\begin{equation}
\chi^2 = (\mu_{\rm obs} - \mu_C)^\dagger C^{-1} (\mu_{\rm obs} - \mu_C)
\end{equation}
where $C^{-1}$ is an uncertainty matrix that combines the uncertainty from the SALT2 fits, intrinsic dispersion, calibration, dust, peculiar velocity and many other factors (see \citet{Betoule2014} for a review). The benefit this analysis methodology provides is speed - for samples of hundreds of supernovae, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with two costs. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of uncertainty Gaussianity. Secondly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae.

Selection efficiency, such as the well known Malmquist bias \citep{MalmquistK.G.1922} is accounted for by correcting data. Simulations following survey observational strategies and geometry are used to calculate the expected bias in distance modulus, which is then added onto the observational data. As these effects are not built into the likelihood, their influence on the error budget is not captured fully in the $\chi^2$ distribution, and any subtle correlations between cosmological or population parameters and the bias is lost.

@@

Sitting comfortably between the traditional models simplicity and the complexity of forward modelling lies Hierarchical Bayesian Models. With the introduction of multiple layers in our model, we can add far more flexibility than a traditional analysis whilst still maintaining most of the computational benefits that come from having a tractable likelihood. \citet{Mandel2009} and \citet{Mandel2011} construct a hierarchical model that they apply to the light curve fitting for supernova. \citet{March2011, March2014, Karpenka2015} derive a hierarchical model and simplify it by analytically marginalising over nuisance parameters to provide a model that offers increased flexibility with reduced uncertainty over the traditional method. The recent BAHAMAS model \citep{Shariff2016} builds off this and reanalyses the JLA dataset, whilst including extra freedom in the correction factors $\alpha$ and $\beta$, finding evidence for redshift dependence on $\beta$. \citet{Ma2016} also performed a reanalysis of the JLA dataset, finding significant differences in $\alpha$ and $\beta$ values from the original as well. Notably, these methods rely on data that is bias corrected, however the UNITY framework given by \citet{Rubin2015} incorporates selection efficiency analytically in the model, and is applied to the Union 2.1 dataset \citep{Suzuki2012}. The well known BEAMS (Bayesian estimation applied to multiple species) methodology from \citet{Kunz2007} has been extended and applied in several works \citep{Hlozek2012}, mostly lately to include redshift uncertainty for photometric redshift application as zBEAMS \citep{Roberts2017} and to include simulated bias corrections in \citet{Kessler2017}. However, whilst there are a large amount of hierarchical models available, none of them have undergone high-statistics simulation verification to quantify each models respective bias, which is becoming critically important as precision supernovae cosmology comes into its own.

The flexibility afforded by a hierarchical model allows for investigations into different treatments of underlying supernova magnitude, colour and stretch populations, supernovae type rates and redshift distributions, host-galaxy corrections and redshift evolution, each of which will be discussed further in the outline of our model below.
